---
title: "Introduction to gtfsio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to gtfsio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The General Transit Feed Specification (GTFS) data format defines a common scheme for describing transit systems, and is widely used by transit agencies around the world and consumed by many software applications. Thus, many **R** packages have been developed to read, write, manipulate and analyse such feeds, such as `{gtfs2gps}`, `{gtfsrouter}`, `{gtfstools}` and `{tidytransit}`.

Each one of these, however, represent GTFS feeds in a slightly different way, making the interoperability between packages harder to accomplish. At the end of the day, this lack of integration results in a more painful experience to the final user who may want to enjoy functions from different packages.

**gtfsio** offers tools for the development of GTFS-related packages that aim to increase such interoperability. It establishes a standard for representing GTFS feeds using R data types. It provides fast and flexible functions to read and write GTFS feeds while sticking to this standard. It defines a basic `gtfs` class which is meant to be extended by packages that depend on it. And it also offers utility functions that support checking the structure of GTFS objects.

This vignette describes the usage of the GTFS reading and writing functions. Please follow the links below for vignettes covering other topics:

- other
- vignettes

# Basic usage

Before using **gtfsio** please make sure that you have it installed in your computer. You can download either the most stable version from CRAN or the development version from GitHub:

```{r, eval = FALSE}
# stable version
install.packages("gtfsio")

# development version
remotes::install_github("r-transit/gtfsio")
```

Then attach it to the current R session:

```{r, message = FALSE}
library(gtfsio)
```

Throughout this demonstration we will be using a few sample files included in the package:

```{r}
data_dir <- system.file("extdata", package = "gtfsio")
list.files(data_dir)
```

- `ggl_gtfs.zip` has been manually built from the [example GTFS feed](https://developers.google.com/transit/gtfs/examples/gtfs-feed) provided by Google. The files samples are licensed under [Creative Commons Attribution 4.0 License](https://creativecommons.org/licenses/by/4.0/).
- `bad_gtfs.zip` is a modified version of `ggl_gtfs.zip` that includes some issues frequently found in GTFS data.
